Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-04-29 concurrency/CONCUR 2023/paper.tex
finished proof of lock-coupling traverse

The definitions of $\nodeboxrep$ and $\treerep$ may vary depending on the data structure, but fundamentally, $\nodeboxrep$ holds the resources owned by each client thread, while $\treerep$ holds the abstract representation of the data structure, treated as an atomically accessed shared resource. In particular, holding $\nodeboxrep(\texttt{pn->n})$ should allow us to acquire the lock on node \texttt{pn->n}, which in turn gives us access to the contents of the node. Formally, we write
\[\nodeboxrep(\texttt{pn->n}) \triangleq\ \texttt{pn->n->lock} \lockvar \mathsf{R}(\texttt{pn->n})\]
where the \emph{lock invariant assertion} $\ell \lockvar \mathsf{R}$ expresses that the lock $\ell$ protects resources $\mathsf{R}$. In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{pn->n}$ and obtain the resources $\mathsf{R}(\texttt{pn->n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. \wm{Should we replace this with a concrete definition of the invariant? e.g. $R = ghost\_node(k, v, range) * ...$}

@@

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{pn->n})$ and the contents $\mathsf{R}(\texttt{pn->n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{pn->n'})$ \wm{This looks weird: \lstinline{pn} doesn't have an \lstinline{n'} field}, its contents $\mathsf{R}(\texttt{pn->n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{pn->n'->t} is \texttt{NULL} or not \wm{Indicate the relationship between \lstinline{res} and the \lstinline{flag} return value}. The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{k} \in \mathsf{range}(\texttt{n'})\ \land \ \texttt{pn} \mapsto (\texttt{p, n}) \ast \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'})
\end{mathpar}\wm{Where does \lstinline{range} come from here?}

\begin{figure}[h]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
	\right\rangle$
	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
		int traverse(pn *pn, int k){
			int flag = 1; 			$\left\{\begin{array}{l}  ...		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
			for( ; ; ){ 			$\left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
				pn->p = pn->n;
				if (pn->p->t == NULL)
				break;
				else{
					int b = findNext(pn, k);
					if (b == 0){
						flag = 0;
						break;
					}
					else{
						acquire(pn->n->lock);
						release(pn->p->lock);
					}
				}
			}
			return flag;
	}  \end{lstlisting}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{n'}) \ \ast \
		\\ \mathsf{R}(\texttt{n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
\end{figure}

Given this invariant, the proof proceeds as follows. \wm{Should we consider giving annotated code after all?} We begin by checking whether the current node is null (line 5 of Figure~\ref{traverse_lock_b}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$. Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{pn->n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$. Otherwise, we acquire \lstinline{pn->n->lock}, gaining access to the resources $\mathsf{R}(\texttt{pn->n})$, and then release the lock of the current node \texttt{pn->p} and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

@@

\subsubsection{Give-up Template}
\label{traverse_proof_giveup}

The atomic specification for the give-up template's \lstinline{traverse} is shown in Figure \ref{fig:traverse_giveup}.

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{c}
				(\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ \infp(\texttt{pn->n'}) 
				\\ 
				\ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

The give-up template's \texttt{traverse} specification is simpler than that of the lock-coupling template. The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{pn->n'})$. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.

%\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly. 
For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held. Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists b. \ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]\wm{Consider using mathsf for logic and reserving texttt for C code.}
%where a lock at location $\ell$ safeguards resources denoted by \texttt{R}. When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$. On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure. As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.

>>>>

The definitions of $\nodeboxrep$ and $\treerep$ may vary depending on the data structure, but fundamentally, $\nodeboxrep$ holds the resources owned by each client thread, while $\treerep$ holds the abstract representation of the data structure, treated as an atomically accessed shared resource. In particular, holding $\nodeboxrep(\texttt{n})$ should allow us to acquire the lock on node \texttt{n}, which in turn gives us access to the contents of the node. Formally, we write
\[\nodeboxrep(\texttt{n}) \triangleq\ \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n})\]
where the \emph{lock invariant assertion} $\ell \lockvar \mathsf{R}$ expresses that the lock $\ell$ protects resources $\mathsf{R}$. In this case, $\mathsf{R}$ contains the contents of the node (i.e., the \lstinline{node} struct contained inside the \lstinline{node_t} struct of \lstinline{pn->n}).

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the resource  $\mathsf{R}(\texttt{n}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{DS\_specific(\texttt{n})}$.

@@

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{n})$ and the contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its boolean state. The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{n'}.\ \texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range}(\texttt{n'})\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \  \mathsf{R}(\texttt{n'}) \ \ast \ \treerep\ m
\end{mathpar} %\wm{Where does \lstinline{range} come from here?}

\begin{figure}[h]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
	\right\rangle$
	\begin{lstlisting}[language = C, style=myStyle, mathescape=true]
		int traverse(pn *pn, int k){
			int flag = 1; 			$\color{blue} \left\{\begin{array}{l}  \mathit{res} = \mathsf{true} \ast  \texttt{pn} \mapsto (\texttt{p, n}) \ast  
				\nodeboxrep(\texttt{n})  \ast \mathsf{R}(\texttt{n})  \ast \treerep\ m 		\end{array}\right\} \Rrightarrow \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\}$
			for( ; ; ){ 			$\color{blue} \left\{\begin{array}{l} \mathsf{traverse\_inv} \end{array}\right\} \triangleq \left\{\begin{array}{l} ... \end{array}\right\}$
				pn->p = pn->n;
				if (pn->p->t == NULL)
				$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} = \texttt{NULL}  \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots \end{array}\right\}$
				break;
				else{
					int b = findNext(pn, k);
					$\color{blue} \left\{\begin{array}{l} \exists \  \texttt{n'}. \ \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \\ 
						(\mathit{res'} = \mathsf{false} \ast \texttt{n'} = \texttt{n} \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \cdots )\  \lor \\ 
						(\mathit{res'} = \mathsf{true}  \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots)   \end{array}\right\}$
					if (b == 0){
						flag = 0;
						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast \texttt{n->t} \neq \texttt{NULL} \ast \texttt{n'} = \texttt{n} \ast \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{k} \in \mathsf{range}(\texttt{n}) \ast \cdots    \end{array}\right\}$
						break;
					}
					else{
						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
						acquire(pn->n->lock);
						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \mathsf{R}(\texttt{n}) \ast \\ \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
						release(pn->p->lock);
						$\color{blue} \left\{\begin{array}{l} \mathit{res} = \mathsf{true} \ast \texttt{pn} \mapsto (\texttt{n, n}) \ast  \texttt{n->lock} \lockvar \mathsf{R}(\texttt{n}) \ast \\ \texttt{n->t} \mapsto \texttt{n'} \ast \texttt{n'->lock} \lockvar \mathsf{R}(\texttt{n'}) \ast \mathsf{R}(\texttt{n'}) \ast \cdots    \end{array}\right\}$
					}
				}
			}
			return flag;
			$\color{blue} \left\{\begin{array}{l} (\mathit{res} = \mathsf{false} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} \neq \texttt{NULL} \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast \texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots) \  \lor \\
				(\mathit{res} =\mathsf{true} \ast \texttt{pn} \mapsto (\texttt{p', p'}) \ast \texttt{p'->t} = \texttt{NULL}  \ast \texttt{p'->lock} \lockvar \mathsf{R}(\texttt{p'}) \ast \mathsf{R}(\texttt{p'}) \ast (\texttt{k} \in \mathsf{range}(\texttt{p'}) \ast \cdots)
			\end{array}\right\}$
	}  \end{lstlisting}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{c} \texttt{pn} \mapsto (\texttt{p, n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \nodeboxrep(\texttt{n'}) \ \ast \
		\\ \mathsf{R}(\texttt{n'})\ \ast \ (\texttt{k} \in \mathsf{range}(\texttt{n'}))
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{The \texttt{traverse} function in lock-coupling template annotated with separation logic specification}
	\label{proof_lock_traverse}
\end{figure}

Given this invariant, the proof proceeds as follows. %\wm{Should we consider giving annotated code after all?} 
We begin by checking whether the current node is null (line 5 of Figure~\ref{proof_lock_traverse}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with $\mathit{res} = \texttt{true}$. Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with $\mathit{res} = \mathtt{false}$. Otherwise, we acquire \lstinline{n->lock}, gaining access to the resources $\mathsf{R}(\texttt{n})$, and then release the lock of the current node \texttt{p} and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish $\mathsf{traverse\_inv}$ for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

@@

\subsubsection{Give-up Proof}
\label{traverse_proof_giveup}

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \infp (\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL}))  \ \ast \ 
				\\ \infp(\texttt{pn->n'}) \ \ast \ \mathsf{R}(\texttt{pn->n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

The give-up template's \texttt{traverse} specification, shown in Figure \ref{fig:traverse_giveup},  is simpler than that of the lock-coupling template. The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents $\mathsf{R}(\texttt{pn->n'})$. Also as before, the Boolean $\mathit{res}$ indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.

%\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly. 
For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held. Formally, we write \[\mathsf{inv\_for\_lock} \ \ell \ \mathsf{R} \triangleq \exists b. \ \ell \mapsto b \ \ast \ \mathsf{if}\ b \ \mathsf{then}\ \mathsf{emp}\ \mathsf{else}\ \mathsf{R}\]
%\wm{Consider using mathsf for logic and reserving texttt for C code.}
%where a lock at location $\ell$ safeguards resources denoted by \texttt{R}. When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$. On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
and $\treerep$ contains an $\mathsf{inv\_for\_lock}$ assertion for each node in the data structure. As before, $\mathsf{R}$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.
